Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature

### 2021-03-11 manuscript/v5_submission_revision/replyreviewer.Rmd

---
title: "reply to reviewer"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*This is an interesting manuscript that aims to identify "causal" relationships" for COVID-19 transmission based on a DAG analysis and empirical data from other studies for variable selection.*

Yes, we aim to estimate causal effects based on a DAG that was constructed based on other publications. Empirical data was extracted and aggregated from different publicly available databases: RKI, Google mobility/Trends, INKAR, DWD. Variable selection for the different models was based on the proposed DAG using the well-studied methods from Pearl etc.

*The DAG analysis is a multivariate analysis to select a subset of variables that are correlated from a large number of candidates. The first select the sets of variables that are associated with the outcome using DAG analysis and then put them into the regression. The DAG analysis returns different sets of variables, using different selection criterion ( for example, the most parsimonious set or the most variation explained ). So the authors assumedly use R-squared to select the final sets of variables in the regression model.*

Yes, we identify subsets of variables using DAG analysis. We stress the point that we do this identification process separately for every variable of interest where we are interested in the causal effect. We always regress on the same outcome Y (Reported COVID-19 cases) using negative-binomial regression. That is, we select one variable Xi where we are interested in the effect of Xi on Y. Then we identify all possible subsets of variables in the DAG that allow to estimate the causal effect of Xi on Y (based on Pearl etc). Out of these subsets we choose the "best" one in regards to pseudo-R2 (or AIC). We repeat this procedure for several variables Xi.

*The challenge I have here (and having reviewed this as well with our statistical team) is that the claims of causal relationships are not convincing, even if they identify associative relationships that other studies have also found. Specifically: 1) The model does not address the nonlinearity of variables (such as temperature); other studies have found that temperature has a u-shaped effect at low and high temperatures. Humidity effects may also be nonlinear. The assumption of non-linearity on continuous variables needs to be considered.*

Yes, non-linearity might be a problem for our generalized linear model (negative-binomial regression). We added a complete check of model diagnostics that addresses linearity assumptions, model specification, independence of observations, and multi-collinearity checks (see below). We did not find significant problems with non-linear relationships (see plot XX supplementary?).

*2) There are no tests for multi-collinearity that are presented, which presents significant concerns for over-fitting the model. Temperature and humidity are an example here. So is the mobility data.*

This an important remark. We added a correlation plot for the continuous variables and assessed variance inflation factors (VIFs) for the full model. We observed high VIFs for the mobility variables and the different intervention variables. We decided to transform the original mobility variables by PCA and used the set of these principal components instead of the original mobility variables. Furthermore, we summarised the intervention variables into one "Intervention" variable (compare new DAG) by summing these binary variables. Analysing the VIFs on the full model with these transformed variables, we are left with only one VIF>5 (PC Mobi 1=5.1xxx), which is a huge improvement on the original model. We want to add that causal graph analysis for regression is more robust to multi-collinearity than traditional regression approaches (cite paper).

*2) Feels like they cherry-pick the variables of interest in the end? Not sure how they arrived at the final variable list for causal effects. Whey were the restrictions not included in that analysis?*

*3) The term “causal analysis” is a bit strong for what they have done here. The basis of the work is the proposed DAG (Figure 1), but the diagram was constructed from other association studies. Other than the DAG analysis, they did not do anything to ensure the results are “causal relationship”. So I am not convinced that the analysis or results are causal.How do you avoid over-fitting the model or including mediating variables in the analysis?*

*5) What about interaction terms? For example, residential mobility and colder weather? Or rain? These relationships are not simply a straightforward multivariable model.*

*6) The most interesting terms in their model were the interventions (restrictions) themselves, yet they were dropped from the model. No discussion is really pursued on that. Why were public health interventions removed?*

### 2021-03-12 manuscript/v5_submission_revision/replyreviewer.Rmd

*2) There are no tests for multi-collinearity that are presented, which presents significant concerns for over-fitting the model. Temperature and humidity are an example here. So is the mobility data.*

This an important remark. We added a correlation plot for the continuous variables and assessed variance inflation factors (VIFs) for the full model. We observed high VIFs for the mobility variables and the different intervention variables. We decided to transform the original mobility variables by PCA and used the set of these principal components instead of the original mobility variables. Furthermore, we summarised the intervention variables into one "Intervention" variable (compare new DAG) by summing these binary variables. Analysing the VIFs on the full model with these transformed variables, we are left with only one VIF>5 (PC Mobi 1=5.1xxx), which is a huge improvement on the original model. We want to add that causal graph analysis for regression is more robust to multi-collinearity than traditional regression approaches (cite paper).

*2) Feels like they cherry-pick the variables of interest in the end? Not sure how they arrived at the final variable list for causal effects. Whey were the restrictions not included in that analysis?*

At the time of writing, we were more interested in variables that could be prognostive of future developments (especially weather, since weather is not modifiable). In light of this remark, we added an overview of the effect estimates derived from the causal graph for all observed variables to the supplementary and discussed them briefly in the main paper.

@@

*3) The term “causal analysis” is a bit strong for what they have done here. The basis of the work is the proposed DAG (Figure 1), but the diagram was constructed from other association studies. Other than the DAG analysis, they did not do anything to ensure the results are “causal relationship”. So I am not convinced that the analysis or results are causal.How do you avoid over-fitting the model or including mediating variables in the analysis?*

We use the term "causal" the same way that Pearl (cite XXX) and the growing body of work based on causal analysis with DAGs use it. We modified the title of our paper to "Causal graph analysis of [...]" to reflect the fundamental assumption of a causal graph to be able to do causal analysis. We are happy to discuss the structure of our proposed DAG.

Proper identification of valid variable subsets for causal effect estimation takes care of the mediating variable problem, since this is at the heart of causal DAG analysis (see Pearl page xxx?).

Overfitting might still be a problem. Since the time of submission the first version of the paper, we were able to "collect" more data of the COVID-19 pandemic. For a sensitivity analysis in regards to overfitting, we re-ran our analysis pipeline on the completely separate dataset from XX-07-2020 to XX-0X-2021 and discuss the corresponding effect estimates in comparison to the results of the original dataset (main paper and supplementary XXX).

@@

*5) What about interaction terms? For example, residential mobility and colder weather? Or rain? These relationships are not simply a straightforward multivariable model.*

DAG analysis explicitly takes care of interaction between variables by identifying valid subsets of variables that need to be included or excluded in the multivariable model, see (cite Pearl some page/chapter?). We added a remark on interaction to the paper (see ...) to reflect on this remark.

@@

*6) The most interesting terms in their model were the interventions (restrictions) themselves, yet they were dropped from the model. No discussion is really pursued on that. Why were public health interventions removed?*

See our answer x) above. We added effect estimates for all variables.

### 2021-03-19 manuscript/v4_medrxiv_revision/causal_model.Rmd
careful wording "causal"

We used a directed acyclic graph (DAG) [@schipf_directed_2011;@textor_robust_2017] as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread. To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar. Restrictions were applied to English and German language and the publication date in the last one year. The following search terms were applied to abstracts and title in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with the exposure variables ("mobility", “public awareness”, "awareness", “google trends”,“ambient temperature”, "temperature"). For "mobility", we analyzed $n=8$ studies, $N=103$ were scanned in Pubmed, together with the first ten pages (100 results) in Google scholar ("awareness"/"public awareness"/"google trends" $n=9$, $N=215$; "temperature"/"ambient temperature" $n=16$, $N=235$). We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \@ref(fig:dag-covid-19).

@@

Google Trends [@google_trends] data can be used as a tool to get insights into public interest (**awareness**) in the coronavirus disease. Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases [@ayyoubzadeh_predicting_2020;@effenberger_association_2020;@higgins_correlations_2020;@li_retrospective_2020;@lin_google_2020;@mavragani_tracking_2020;@walker_use_2020;@yuan_trends_2020;@zhou_effects_2020]. Some search terms e.g. "COVID-19" or "coronavirus" predated newly infected cases/total number of cases by roughly 7 to 14 days for different countries [@effenberger_association_2020; @higgins_correlations_2020;@li_retrospective_2020;@yuan_trends_2020]. Additionally, we acknowledged that individual risk-aware behavior might be a reaction to the current COVID-19 burden (measured as reported cases at the day of exposure).

@@

When investigating causal determinants of SARS-CoV-2 infections, a number of confounders have to be considered. Well-known risk factors for SARS-CoV-2 as well as for other infections are demographic factors such as age, gender, socio-economic status (SES), population density, and foreign citizenship/ethnicity [@de2020risk;@Dragano2020.06.17.20133918;@jhucovid19db2020]. In Germany along with other countries (i.e. Brazil, USA, or the UK), populist parties or politicians and their electorate tend to be more sceptical about effects of containment measures than the other part of the electorate [@dohle_wingen_schreiber_2020;@engle_staying_2020]. Therefore we considered both "right-wing populist party votes" and "voter turnout" as possible confounders. Public health interventions were also taken into account (contact restrictions, school closures etc.), as their implementation showed strong correlations with controlling the spread of SARS-CoV-2 [@cowling2020impact;@juni_impact_2020;@lai_effect_2020]. To avoid bias due to reporting delay of case numbers we had to include weekday and German holidays. We include some unobserved variables in our DAG (e.g. "Herd immunity"), too. Please note that "Exposure to SARS-CoV-2" is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset. *Exposure to the virus* should not be confused with the formal *exposure variables* of the DAG (mobility, awareness, temperature).

>>>>

We used a directed acyclic graph (DAG) [@schipf_directed_2011;@textor_robust_2017] as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread. To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar. Restrictions were applied to English and German language and the publication date in the last one year. The following search terms were applied to abstracts and title in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with a selection of exposure variables ("mobility", “public awareness”, "awareness", “google trends”, “ambient temperature”, "temperature"). For "mobility", we analyzed $n=8$ studies, $N=103$ were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar ("awareness"/"public awareness"/"google trends" $n=9$, $N=215$; "temperature"/"ambient temperature" $n=16$, $N=235$). We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \@ref(fig:dag-covid-19).

@@

Google Trends [@google_trends] data can be used as a tool to get insights into public interest (**awareness**) in COVID-19. Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases [@ayyoubzadeh_predicting_2020;@effenberger_association_2020;@higgins_correlations_2020;@li_retrospective_2020;@lin_google_2020;@mavragani_tracking_2020;@walker_use_2020;@yuan_trends_2020;@zhou_effects_2020]. Some search terms e.g. "COVID-19" or "coronavirus" predated newly infected cases/total number of cases by roughly 7 to 14 days for different countries [@effenberger_association_2020; @higgins_correlations_2020;@li_retrospective_2020;@yuan_trends_2020]. Additionally, we acknowledged that individual risk-aware behavior might be a reaction to the current COVID-19 burden (measured as reported cases at the day of exposure).

@@

When investigating causal determinants of SARS-CoV-2 infections, a number of confounders have to be considered. Well-known risk factors for SARS-CoV-2 as well as for other infections are demographic factors such as age, gender, socio-economic status (SES), population density, and foreign citizenship/ethnicity [@de2020risk;@Dragano2020.06.17.20133918;@jhucovid19db2020]. In Germany along with other countries (i.e. Brazil, USA, or the UK), populist parties or politicians and their electorate tend to be more sceptical about effects of containment measures than the other part of the electorate [@dohle_wingen_schreiber_2020;@engle_staying_2020]. Therefore we considered both "right-wing populist party votes" and "voter turnout" as possible confounders. Public health interventions were also taken into account (contact restrictions, school closures etc.), as their implementation showed strong correlations with controlling the spread of SARS-CoV-2 [@cowling2020impact;@juni_impact_2020;@lai_effect_2020]. To avoid bias due to reporting delay of case numbers we had to include weekday and German holidays. We include some unobserved variables in our DAG (e.g. "Herd immunity"), too. Please note that "Exposure to SARS-CoV-2" is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset. *Exposure to the virus* should not be confused with the formal *exposure variables* of the DAG.
